Enforced Invariant Confluence

### 2020-01-14 doc/the_vldb_journal_paper/sections/segmented_invariant_confluence.tex
Added tech report sections of paper.
Revised Segmented Invariant Confluence section.

If a distributed object is \invariantconfluent{}, then the object can be
replicated without the need for any form of coordination to maintain the
object's invariant. But what if the object is \emph{not} \invariantconfluent{}?
In this section, we present a generalization of \invariantconfluence{} called
\defword{segmented \invariantconfluence{}} that can be used to maintain the
invariants of non-\invariantconfluent{} objects, requiring only a small amount
of coordination. In \secref{Evaluation}, we see that replicating a
non-\invariantconfluent{} object with segmented \invariantconfluence{} can
achieve between 10x and 100x more throughput than linearizable replication for
certain workloads.

@@

\begin{example}\examplelabel{SegmentedZ2}
  Consider again the object $O = (\ints \times \ints, \join)$, transactions $T
  = \set{t_{x+1}, t_{y-1}}$, and invariant $I = \setst{(x, y)}{xy \leq 0}$ from
  \exampleref{Z2}, but now let the start state $s_0$ be $(-42, 42)$. $O$ is
  \emph{not} \sTIconfluent{} because the points $(0, 42)$ and $(42, 0)$ are
  reachable, and merging these points yields $(42, 42)$ which violates the
  invariant. However, $O$ is \sTISconfluent{} for $\Sigma = (I_1, T_1)$, $(I_2,
  T_2)$, $(I_3, T_3)$, $(I_4, T_4)$ where
  \begin{align*}
    I_1 &= \setst{(x, y)}{x < 0, y > 0}       &T_1 &= \set{t_{x+1}, t_{y-1}} \\
    I_2 &= \setst{(x, y)}{x \geq 0, y \leq 0} &T_2 &= \set{t_{x+1}, t_{y-1}} \\
    I_3 &= \setst{(x, y)}{x = 0}              &T_3 &= \set{t_{y-1}} \\
    I_4 &= \setst{(x, y)}{y = 0}              &T_4 &= \set{t_{x+1}}
  \end{align*}
  $\Sigma$ is illustrated in \figref{SegmentedZ2}. Clearly, $s_0$ satisfies the
  invariant, and $I_1, I_2, I_3, I_4$ cover $I$. Moreover, for every $(I_i,
  T_i) \in \Sigma$, we see that $O$ is \iclosed{I_i}, so $O$ is
  \sticonfluent{s}{T_i}{I_I} for every $s \in I_i$. Thus, $O$ is
  \sTISconfluent{}.
\end{example}

@@

\subsection{System Model}
Now, we describe the system model used to replicate a segmented
\invariantconfluent{} object without any coordination within a segment and with
only a small amount of coordination when transitioning between segments. As
before, we replicate an object $O$ across a set $p_1, \ldots, p_n$ of $n$
servers each of which manages a replica $s_i \in S$ of the object. Every server
begins with $s_0$, $T$, $I$, and $\Sigma$. Moreover, at any given point in
time, a server designates one of the segments in $\Sigma$ as its
\defword{active segment}.  Initially, every server chooses the first segment
$(I_i, T_i) \in \Sigma$ such that $I_i(s_0)$ to be its active segment. We'll
see momentarily the significance of the active segment.

As before, servers repeatedly perform one of two actions: execute a transaction
or merge states with another server. Merging states in the segmented
\invariantconfluence{} system model is identical to merging states in the
\invariantconfluence{} system model. A server $p_i$ sends its state $s_i$ to
another server $p_j$ which \emph{must} merge $s_i$ into its state $s_j$.
%
Transaction execution in the new system model, on the other hand, is a bit more
involved. Consider a server $s_i$ with active segment $(I_i, T_i)$. A client
can request that $p_i$ execute a transaction $t \in T$. We consider what
happens when $t \in T_i$ and $t \notin T_i$ separately.

@@

\begin{algorithm}[t]
  \caption{%
    Transaction execution in the segmented \invariantconfluence{} system model}%
  \algolabel{TxnExecution}
  \begin{algorithmic}
    \If{$t \notin T_i$}
      \State Execute $t$ with global coordination
    \Else{}
      \If{$I_i(t(s_i))$} Commit $t$
      \ElsIf{$\lnot I(t(s_i))$} Abort $t$
      \Else{} Execute $t$ with global coordination
      \EndIf{}
    \EndIf{}
  \end{algorithmic}
\end{algorithm}

@@

\begin{theorem}\thmlabel{ExtendedIclosureEquivalentIconfluence}
  Consider an object $O = (S, \join)$, a set of transactions $T$, and an
  invariant $I$. If every pair of states in the invariant are \TIcoreachable{},
  then
  \[
    \text{$O$ is \Iclosed{}}
    \iff
    \text{$O$ is \sticonfluent{s}{T}{I} for every $s \in I$}
  \]
\end{theorem}

The proof of the forward direction is exactly the same as the proof of
\thmref{IclosureImpliesIconfluence}. Transactions always maintain the
invariant, so if merge does as well, then every reachable state must satisfy
the invariant.
%
For the reverse direction, consider two arbitrary states $s_1, s_2 \in I$.
The two points are \TIcoreachable{}, so there exists some state $s_0$ from
which they can be reached. $O$ is \sTIconfluent{} and $s_1 \join s_2$ is
\sTIreachable{}, so it satisfies the invariant.

@@

\begin{techreport}

@@

  Fourth, we can integrate a couple of optimizations into our system model to
  further reduce the amount of coordination it requires. To begin, if a server
  with state $s_i$ and active segment $(I_i, T_i)$ receives a transaction $t
  \in I_i$ to execute, and $t(s_i)$ violates the active invariant but not the
  global invariant, instead of initiating a round of global coordination, $p_i$
  can simply buffer $t$ for re-execution at a later time. While this increases
  the latency required to execute $t$, it's possible that after other
  transactions are executed, re-executing $t$ may lead to a state that either
  satisfies the active invariant or violates the global invariant. In either
  case, a round of global coordination is avoided. Similarly, servers can
  buffer transactions that require global coordination, executing an entire
  batch of these transactions during a single round of global coordination.

  Fifth, a segmented \invariantconfluence{} decision procedure can also
  leverage \thmref{LatticeProperty} in addition to
  \algoref{ArbitraryStartInteractiveDecisionProcedure}. If an object $O$ meets
  criteria (1) - (3), then it is \sticonfluent{s}{T}{I} for every state $s \in
  I$.
\end{techreport}

>>>>

If a distributed object is \invariantconfluent{}, then the object can be
replicated without the need for any form of coordination to maintain the
object's invariant. But what if the object is \emph{not} \invariantconfluent{}?
In this section, we present a generalization of \invariantconfluence{} called
\defword{segmented \invariantconfluence{}} that can be used to maintain the
invariants of non-\invariantconfluent{} objects, requiring only a small amount
of coordination. In \secref{Evaluation}, we see that replicating a
non-\invariantconfluent{} object with segmented \invariantconfluence{} can
achieve between 10$\times$ and 100$\times$ more throughput than linearizable
replication for certain workloads.

@@

\begin{example}\examplelabel{SegmentedZ2}
  Consider again the object $O = (\ints \times \ints, \join)$, the set of
  transactions $T = \set{t_{x+1}, t_{y-1}}$, and the invariant $I = \setst{(x,
  y)}{xy \leq 0}$ from \exampleref{Z2}, but now let the start state $s_0$ be
  $(-42, 42)$. $O$ is \emph{not} \sTIconfluent{} because the points $(0, 42)$
  and $(42, 0)$ are reachable, and merging these points yields $(42, 42)$ which
  violates the invariant. However, $O$ is \sTISconfluent{} for $\Sigma = (I_1,
  T_1)$, $(I_2, T_2)$, $(I_3, T_3)$, $(I_4, T_4)$ where
  \begin{align*}
    \qquad I_1 &= \setst{(x, y)}{x < 0, y > 0}       &T_1 &= \set{t_{x+1}, t_{y-1}} \\
    \qquad I_2 &= \setst{(x, y)}{x \geq 0, y \leq 0} &T_2 &= \set{t_{x+1}, t_{y-1}} \\
    \qquad I_3 &= \setst{(x, y)}{x = 0}              &T_3 &= \set{t_{y-1}} \\
    \qquad I_4 &= \setst{(x, y)}{y = 0}              &T_4 &= \set{t_{x+1}}
  \end{align*}
  $\Sigma$ is illustrated in \figref{SegmentedZ2}. Clearly, $s_0$ satisfies the
  invariant, and $I_1, I_2, I_3, I_4$ cover $I$. Moreover, for every $(I_i,
  T_i) \in \Sigma$, we see that $O$ is \iclosed{I_i}, so $O$ is
  \sticonfluent{s}{T_i}{I_I} for every $s \in I_i$. Thus, $O$ is
  \sTISconfluent{}.
\end{example}

@@

\subsection{System Model}
Now, we describe the system model used to replicate a segmented
\invariantconfluent{} object without any coordination within a segment and with
only a small amount of coordination when transitioning between segments. As
before, we replicate an object $O$ across a set $p_1, \ldots, p_n$ of $n$
servers each of which manages a replica $s_i \in S$ of the object. Every server
begins with $s_0$, $T$, $I$, and $\Sigma$. Moreover, at any given point in
time, a server designates one of the segments in $\Sigma$ as its
\defword{active segment}.  Initially, every server selects its active segment
to be the first segment $(I_i, T_i) \in \Sigma$ such that $I_i(s_0)$. We'll see
momentarily the significance of the active segment.

As before, servers repeatedly perform one of two actions: execute a transaction
or merge states with another server. Merging states in the segmented
\invariantconfluence{} system model is identical to merging states in the
\invariantconfluence{} system model. A server $p_i$ sends its state $s_i$ to
another server $p_j$ which \emph{must} merge $s_i$ into its state $s_j$.
%
Transaction execution in the new system model, on the other hand, is a bit more
involved. Consider a server $p_i$ with active segment $(I_i, T_i)$. A client
can request that $p_i$ execute a transaction $t \in T$. We consider what
happens when $t \in T_i$ and $t \notin T_i$ separately.

@@

\begin{algorithm}[t]
  \caption{%
    Transaction execution in the segmented \invariantconfluence{} system model}%
  \algolabel{TxnExecution}
  \begin{algorithmic}
    \If{$t \notin T_i$}
      \State Execute $t$ with global coordination
    \Else{}
      \If{$I_i(t(s_i))$}
        \State Commit $t$
      \ElsIf{$\lnot I(t(s_i))$}
        \State Abort $t$
      \Else{}
        \State Execute $t$ with global coordination
      \EndIf{}
    \EndIf{}
  \end{algorithmic}
\end{algorithm}

@@

\begin{example}
  Consider the object $O = (\nats, \max)$ consisting of natural numbers merged
  by the max function. Let $I = \nats$. Let $T = \set{t_{2\times}, t_{+10}}$
  where $t_{2\times}(x) = 2x$ and $t_{+10} = x + 10$. The states $3$ and $16$
  are \TIcoreachable{} because both are \stireachable{s}{T}{I} from $s = 3$
  ($16 = 2 \times 3 + 10$). The states $3$ and $5$, however, are not
  \TIcoreachable{}. The only state in which $3$ is reachable is $3$, but from
  this state, the smallest reachable number larger than $3$ is $6$.
\end{example}

\begin{theorem}\thmlabel{ExtendedIclosureEquivalentIconfluence}
  Consider an object $O = (S, \join)$, a set of transactions $T$, and an
  invariant $I$. If every pair of states in the invariant are \TIcoreachable{},
  then
  \[
    \hfill
    \text{$O$ is \Iclosed{}}
    \iff
    \forall s \in I.\ \text{$O$ is \sticonfluent{s}{T}{I}}
    \hfill
  \]
\end{theorem}

\begin{proof}
  The proof of the forward direction is exactly the same as the proof of
  \thmref{IclosureImpliesIconfluence}. Transactions always maintain the
  invariant, so if merge does as well, then every reachable state must satisfy
  the invariant.
  %
  For the reverse direction, consider two arbitrary states $s_1, s_2 \in I$.
  By assumption, the two points are \TIcoreachable{}, so there exists some
  state $s_0$ from which they can be reached. $O$ is \sTIconfluent{} and $s_1
  \join s_2$ is \sTIreachable{}, so it satisfies the invariant.
\end{proof}

@@

@@

Fourth, we can integrate a couple of optimizations into our system model to
further reduce the amount of coordination it requires. To begin, if a server
with state $s_i$ and active segment $(I_i, T_i)$ receives a transaction $t
\in T_i$ to execute, and $t(s_i)$ violates the active invariant but not the
global invariant, instead of initiating a round of global coordination, $p_i$
can simply buffer $t$ for re-execution at a later time. While this increases
the latency required to execute $t$, it's possible that after other
transactions are executed, re-executing $t$ may lead to a state that either
satisfies the active invariant or violates the global invariant. In either
case, a round of global coordination is avoided. Similarly, servers can
buffer transactions that require global coordination, executing an entire
batch of these transactions during a single round of global coordination.

Fifth, a segmented \invariantconfluence{} decision procedure can also
leverage \thmref{LatticeProperty} in addition to
\algoref{ArbitraryStartInteractiveDecisionProcedure}. If an object $O$ meets
criteria (1) - (3), then it is \sticonfluent{s}{T}{I} for every state $s \in
I$.

Sixth, a naive implementation of our segmented invariant confluence system
model is \emph{not} fault tolerant. A round of global coordination requires
that \emph{every} server be non-faulty. If even a single server fails, then
every round of global coordination will fail to complete. Fortunately, we can
leverage well-established means of ensuring fault tolerance without difficulty.
For example, we can replicate every server using a state machine replication
protocol like MultiPaxos~\cite{lamport1998part, lamport2001paxos} or
Raft~\cite{ongaro2014search}. Doing so will introduce a constant slowdown for
each (now replicated) server, but servers remain independent and scalable.
